A performance comparison of Dask and Spark for data-intensive neuroimaging pipelines

### 2019-07-24 latex/paper.tex
Edited the introduction

\title{A Comparative study of Dask and Spark for data-intensive neuroimaging pipelines}

@@

\section{Introduction}
% Context
\VHS{it's not the the processing of BD becomes critical, it's that the results obtained
from processing more data should be more reliable\dots and the data is there too}
The recent rise in data sharing and improved data collection strategies
have brought neuroimaging to the Big Data era ~\cite{ALFAROALMAGRO:18,
UKBioBank:18}. Existing neuroimaging frameworks, such as Nipype~\cite{Nipype:11},
are well suited for processing the standard compute-intensive neuroimaging pipelines,
but lack incorporation Big Data strategies (i.e. in-memory computing,
data locality and lazy-evaluation) to improve performance of the increasingly
prevalent data-intensive pipelines. As was noted in ~\cite{Hayot-Sasson:17}, 
in-memory computing, coupled with data locality, can bring significant performance
improvements to data intensive neuroimaging pipelines. We extend this work by studying
the differences between two Big Data frameworks, Dask~\cite{Dask:15} and Apache Spark~\cite{Spark:16},
for their suitability in the processing neuroimaging pipelines.

%Similarities
\TG{Note for later: explain the following concepts in the intro: lazy eval, data locality, in-memory computing.}
Spark and Dask offer in-memory computing, data locality, and lazy evaluation; which
is common for Big Data framework. Both their scheduler operates dynamically. This is
good when the runtimes are not known ahead of time~\cite{Dask:15}. Over these
similarities, the frameworks are quite different.

% Spark
On the one hand, we have Spark which provides a high-level API. This allows the
scheduler to perform more optimizations which makes it well suited for neuroimaging
analysis that often requires the usage of a pipeline with multiple steps. Though,
Spark's code base is in Scala which theoretically can lead to slow down in execution
due to a required serialization. Moreover, while Spark's API is flexible and allows
most implementations, it differs from the ones seen in the Python's ecosystem.

% Dask
On the other hand, Dask was created with the purpose of natively parallelize Python
pipelines while keeping the syntax of familiar API from the Python's scientific
ecosystem. However, Dask is still a young framework with work to be done; it's API
does not completely replicate the library it supports. While its lower-level API
allows the implementation of more complex algorithms it sacrifices a layer of
optimization.
% Related work
Previous work shows that Dask had significant overhead and was
hard to debug~\cite{Mehta:17}.
% What issues your work addresses
Dask was immature at the time and a lot of change was brought to the framework.
Therefore we think it is valuable to re-compare it with Spark.
\TG{Cite~\cite{hayot2019performance}}

%% Dask API
The Dask APIs we decide to use for our comparison are Dask Bag, Delayed and Futures.
Dask Bag offers an easy API to parallelize data. Dask Delayed offer a lower-level API
that offers more flexibility; this is good to implement more complex tasks that do
not fit in the Dask Bag framework. Dask Futures is a real-time API. Both Dask Bag and
Delayed apply lazy evaluation to tasks while Futures trigger them directly.

% Methods used (summary)
The project aims to compare the state-of-the-art general-purpose framework Spark with
the newcomer Dask. We decide to compare the performance of Spark and Dask on a custom
incrementation pipeline to simplify the effect of the algorithm on the comparison.
Then we assess the frameworks on two real-life applications: (1) histogram of the
voxels intensity in a 3d images (2) BIDS example.

% Implications of our research
The result from our project help in deciding if Dask is a good choice to build
neuroimaging pipelines.

\TG{Say that it is often said that Spark has a larger overhead but scales 
worse than Dask. In the dicussion, get back to it and mention that (1) we don't
always observe that, (2) for data-intensive applications, overheads don't really matter
as reducing them increases data trasnfers.}

@@

\subsubsection{Apache Spark} Apache Spark is a general-purpose Big Data engine.
Spark's core is written in Scala, but it also has a Java, an R, and a Python API
called PySpark. We used PySpark, given the popularity of Python in neuroscience,
although the serialization cost of Python objects to Scala might be a substantial
source of overhead. Spark provides in-memory computing, data locality, and lazy
evaluation. PySpark's fundamental data structure, the Resilient Distributed Dataset
(RDD)~\cite{RDD}, is a fault-tolerant and a parallel collection of data elements.
Dataframes, PySpark's API for tabular data, is an overlay of RDDs with similar
performance features. We used Spark's default scheduler, Spark standalone, although
YARN~\cite{vavilapalli2013apache} and Mesos~\cite{hindman2011mesos} are also
available. In the Spark standalone scheduler, a process called the \emph{master}
coordinates resources provisioned by \emph{worker} processes on the cluster. The
application is submitted to the \emph{driver} that in turn requests workers to the
master and dispatches tasks to them. A job is divided into stages to be excuted in a
different process onto the workers. Each stage's operation is represented as a
high-level task in the computation graph. The Spark standalone scheduler uses a FIFO
(First-In-First-Out) job scheduling policy. Processing jobs in order helps in
increasing the page hit rate. As our applications only have one job, the LIFO
(Last-In-First-Out) policy used for the stages is more pertinent \MD{Find better
phrasing} \TG{refer to the paper (\MD{Not sure which one})} \VHS{I wouldn't
explicitly state LIFO here. It's a bit complicated to be frank. Basically both Spark
and Dask opt to complete a chain of dependencies rather than start a bunch of ``new
chains''. And in this context, LIFO is great for the page hit rate, so you might want
to reword the prior sentence as well}. The Spark standalone scheduler has two
execution modes: client mode, where the driver runs in a dedicated process on the
cluster, and cluster mode, where the driver runs in a worker. We used client mode as
cluster mode is not available in PySpark. We used Apache Spark v2.4.0.

@@

In Dask, a \href{https://docs.dask.org/en/latest/bag.html}{Bag} is a parallel
collection of Python objects, similar to Spark's RDD. It use multi-process to enable
parallelism. It offers a programming abstraction similar to the
\href{https://toolz.readthedocs.io/en/latest/}{PyToolz library}. An
\href{https://docs.dask.org/en/latest/array.html}{Array} is used for the processing
of large arrays. It provides a distributed clone of the popular NumPy library. A
\href{https://docs.dask.org/en/latest/dataframe.html}{Dataframe} is a parallel
composition of
\href{http://pandas.pydata.org/pandas-docs/stable/reference/api/pandas.DataFrame.html}{Pandas
Dataframes} used to process a large amount of tabular data.
\href{https://docs.dask.org/en/latest/delayed.html}{Delayed} supports arbitrary tasks
that do not fit in the Array, DataFrame or Bag APIs. Finally,
\href{https://docs.dask.org/en/latest/futures.html}{Futures} are similar to Delayed
as they support arbitrary tasks, but they operate in real-time rather than lazily.

@@

While Incrementation and Histogram were implemented directly in Python, BIDS app
example requires an external library and is distributed as a Docker container image
(bids/base\_fsl on DockerHub). We converted the Docker image to a Singularity image,
given its popularity for HPC and reproducibility, using
\href{https://hub.docker.com/r/singularityware/docker2singularity/tags/}{docker2singularity}.
The image was preloaded on the nfs and singularity version 3.2.1\-1.el7 was installed
on all instances. Due to a bug in the conversion from Docker to Singularity, we had to
export the path for fsl.

\subsection{Experiments}

@@

For each run, we measured the application makespan as well as the cumulative 
data read, compute, data write, and engine overhead across all application tasks. 
We measured the overhead by \TG{\ldots, maybe add a figure}. \MD{Done but might need
adjustment depending on how the overhead is calculate (idle vs wasted)}

@@

%%% BIDS EXAMPLE %%%
\subsection{BidsApp example: Number of workers}
Figure~\ref{fig:bids_ms_worker} shows the makespan of the application while varying
the amount of worker. Overall, there is no substantial difference between the
engines however Futures are always slightly faster. The performance difference between
Futures and the other APIs increases with the number of workers. The makespan scales
well with the increase in workers however the scaling reduces considerably when
reaching 8 workers. This is due to the IO requirement increasing with the number of
workers, since more tasks are read concurrently, and the application becomes more
data-intensive.

@@

>>>>

\title{A performance comparison of Dask and Spark for data-intensive neuroimaging pipelines}

@@

\section{Introduction}
% Context
The recent rise in data sharing and improved data collection strategies
have brought neuroimaging to the Big Data era ~\cite{ALFAROALMAGRO:18,
UKBioBank:18}. Existing neuroimaging workflow engines, such as
Nipype~\cite{Nipype:11}, are well suited for processing the standard
compute-intensive neuroimaging pipelines, but lack incorporated Big Data
strategies (i.e. in-memory computing, data locality and lazy-evaluation) to
improve performance of the increasingly prevalent data-intensive pipelines.
As was noted in ~\cite{hayot2019performance}, in-memory computing, coupled
with data locality, can bring significant performance improvements to data
intensive neuroimaging pipelines. We extend this work by studying the
differences between two Big Data frameworks, Dask~\cite{Dask:15} and Apache
Spark~\cite{Spark:16}, for their suitability in the processing of neuroimaging
pipelines. Our goal is to test whether Spark or Dask has a clear
performance advantage to process Big neuroimaging Data.

%Similarities
Spark and Dask both offer in-memory computing, data locality, and lazy
evaluation; which is common for Big Data frameworks. Both their schedulers
operate dynamically, which is good when task runtimes are not known ahead of
time~\cite{Dask:15}. They also provide rich, high-level programming APIs,
and support a variety of infrastructure schedulers, such as Mesos (Spark
only), YARN, or HPC clusters (Dask only). Over these similarities, the
frameworks have differences.

% Differences
First and foremost, Spark is written in Scala while Dask is in Python.
Given the popularity of Python in scientific communities, this arguably
gives an edge to Dask due to data serialization costs from Python to Scala.
On the other hand, Python's Global Interpreter Lock (GIL) might reduce
parallelism in some cases. This difference in programming languages also
has qualitative implications. As part of the \href{http://scipy.org}{SciPy}
ecosystem, Dask provides almost transparent parallelization of applications
manipulating Numpy arrays or Pandas data frames. On the other hand, Spark's
Java, R and Python APIs allow to easily parallelize analyses that combine
these languages, with reduced performance loss.

% Related work
Spark and Dask were both included in the evaluation reported
in~\cite{Mehta:17} that used a neuroimaging and an astronomy application
processing approximately 100~GB of data each. In this work, Dask was
reported to have a slight performance advantage over Spark. Overall, Dask's
end-to-end time (makespan) was measured to be up to 14\% faster than Spark,
due to ``more efficient pipelining'' %(Fig 10) 
and serialization time to
Python. % (Fig. 12-a)
Dask, however, was reported to have a larger startup
time than Spark. % (Fig. 12-c)
The analysis remains at a quite high level
though, leaving most of the observed performance difference unexplained. In
comparison, our study will provide a detailed analysis of performance
differences and similarities between Spark and Dask.

The next section details the design of our experiments. We consider two
data-intensive neuroimaging applications: high-resolution imaging,
represented by the Big Brain data~\cite{Amunts:13}, and large functional
MRI studies, represented by data from the consortium for reliability and
reproducibility (CoRR~\cite{zuo2014open}). We test application pipelines
involving different patterns (map-only, map-reduce), and different types of
implementations (plain Python, command-line,
containerized). We evaluate performance on a dedicated cluster
representative of the ones used in today's data-intensive neuroimaging
studies, using the main Dask and Spark APIs. The following sections present
our results, discussion and conclusions.

@@

\subsubsection{Apache Spark} Apache Spark is a general-purpose Big Data engine. It
provides in-memory computing, data locality, and lazy evaluation. Spark's main data
structure, the Resilient Distributed Dataset (RDD)~\cite{RDD}, is a fault-tolerant
and a parallel collection of data elements. Dataframes, Spark's API for tabular data,
is an overlay of RDDs with similar performance features \VHS{It's probably fine, but
the SparkSQL Engine available for Scala and Java do bring an added peformance
value}.\VHS{Spark has another API (Datasets). i know you used 'popular' for
DataFrames, but it's not like Spark has a plethora of APIs. I think the APIs should
be mentioned after the API languages such that you only list the APIs relevant to
Python (RDD and Dataframes)} Spark's core is written in Scala, but it also has a
Java, an R, and a Python API called PySpark. \MD{Sentence is ambiguous? could think
that all three APIs are named PySpark} We used PySpark, given the popularity of
Python in neuroscience, although the serialization cost of Python objects to Scala
might be a substantial source of overhead. We used Spark's default scheduler, Spark
standalone, although YARN~\cite{vavilapalli2013apache} and
Mesos~\cite{hindman2011mesos} are also available. In the Spark standalone scheduler,
a process called the \emph{master} coordinates resources provisioned by
\emph{workers} on the cluster. The application is submitted to the \emph{driver} that
in turn requests workers to the master and dispatches tasks to them. A job is divided
into stages to be excuted in a different process onto the workers. Each stage's
operation is represented as a high-level task in the computation graph. The Spark
standalone scheduler uses a FIFO (First-In-First-Out) job scheduling policy.
Processing jobs in order helps in increasing the page hit rate. As our applications
only have one job, the LIFO (Last-In-First-Out) policy used for the stages is more
pertinent \MD{Find better phrasing} \TG{refer to the paper (\MD{Not sure which one})}
\VHS{I wouldn't explicitly state LIFO here. It's a bit complicated to be frank.
Basically both Spark and Dask opt to complete a chain of dependencies rather than
start a bunch of ``new chains''. And in this context, LIFO is great for the page hit
rate, so you might want to reword the prior sentence as well}. The Spark standalone
scheduler has two execution modes: client mode, where the driver runs in a dedicated
process on the cluster, and cluster mode, where the driver runs in a worker. We used
client mode as cluster mode is not available in PySpark. We used Apache Spark v2.4.0.

@@

In Dask, a \href{https://docs.dask.org/en/latest/bag.html}{Bag} is a parallel
collection of Python objects, similar to Spark's RDD. It use multi-process to enable
parallelism. It offers a programming abstraction similar to the
\href{https://toolz.readthedocs.io/en/latest/}{PyToolz library}. An
\href{https://docs.dask.org/en/latest/array.html}{Array} is used for the processing
of large arrays. It provides a distributed clone of the popular NumPy library. A
\href{https://docs.dask.org/en/latest/dataframe.html}{Dataframe} is a parallel
composition of
\href{http://pandas.pydata.org/pandas-docs/stable/reference/api/pandas.DataFrame.html}{Pandas
Dataframes} used to process a large amount of tabular data.
\href{https://docs.dask.org/en/latest/delayed.html}{Delayed} supports arbitrary tasks
that do not fit in the Array, DataFrame or Bag APIs. Finally,
\href{https://docs.dask.org/en/latest/futures.html}{Futures} are similar to Delayed
as they support arbitrary tasks, but they operate in real-time rather than lazily.
\VHS{maybe mention which one use threads and which use processes}

@@

While Incrementation and Histogram were implemented directly in Python,
BIDS app example requires an external library and is distributed as a
Docker container image (bids/base\_fsl on DockerHub). To avoid remote file
transfers from DockerHub, we preloaded our OpenStack instances with the
Docker image, and started the Docker engine at startup.
\subsection{Experiments}

@@

For each run, we measured the application makespan as well as the cumulative 
data read, compute, data write, and engine overhead across all application tasks. 
We measured the overhead by \TG{\ldots, maybe add a figure}.

@@

%%% BIDS EXAMPLE %%%
\subsection{BidsApp example: Number of workers}
Figure~\ref{fig:bids_ms_worker} shows the makespan of the application while varying
the amount of worker. Overall, there is no substantial difference between the
engines however Futures are always slightly faster. The performance difference between
Futures and the other APIs increase with the number of workers. The makespan scales
well with the increase in workers however the scaling reduces considerably when
reaching 8 workers. This due to the IO requirement increasing with the number of
workers, since more tasks are read concurrently, and the application becomes more data
intensive.

@@

% partitioning issue
